Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-05-05 ms/ms.Rnw
further tweaks -- still not really doing results/discussion!

\newcommand{\shir}{Shirreff et al.}%{\citeauthor{shirreff_transmission_2011}}
\newcommand{\champr}{Champredon et al.}%{\citeauthor{champredon_hiv_2013}}
\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
\newcommand{\rzero}{{\cal R}_0}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=small]{todonotes}

@@

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. The trade-off theory \citep{Ebert1999} --- which
postulates that parasite virulence can be explained as the long-term
evolutionary outcome of a saturating relationship between parasite
clearance rate and transmission rate --- has been criticized
\citep{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in several important cases \citep{Dwyer+1990}
\todo{others? de Roode and Altizer? maybe \cite{mackinnon1999genetic},
  \cite{jensen2006empirical}, and \cite{de2008virulence}?}. One
particularly interesting application of these ideas is the work by
Fraser et al. showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in a study of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
within-couple transmission
\citep{Fraser+2007,fraser_virulence_2014}\todo{read/check fraser+
  2014!; check vs. \cite{herbeck_is_2012}}. Subsequent studies
\citep{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately [XXX] years and then declining
slightly to a long-stable virulence level.

@@

\begin{itemize}
\item Since we assume that SPVL is determined by genoyype, we should probably mention papers like this:\cite{alizon2010phylogenetic}
\item Do we need to mention things like heritability of SPVL or do we not have to mention it since it's already included in Shirreff et al's paper and our models are based on their model? It's an important assumption and maybe we need to cite source..?
\end{itemize}

\subsection{Model formulation}

@@

\todo{was going to call this ``within-host'' but it includes transmission}

@@

In contrast, models 5 and 6 are not explicitly structured.  Model 5 is an implicit serial monogamy model based on the epidemiological model used by \shir. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008} \todo{is this the right ref? where does Shirreff's model originate?} \todo{Evolutionary dynamic (SPVL distribution, disease progression rate, mutation, and etc) comes from \cite{Fraser+2007} and serial monogamy approximation comes from \cite{hollingsworth_hiv1_2008}}. Finally, model 6 is a simple random-mixing model.

@@

\begin{equation}
z_i = \text{exp}(\log(z_{min}) + [\log(z_{max}) - \log(z_{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
\end{equation}
For simplicity, we assume that all parameters follow logarithmic distribution. Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$). Then, each column is replaced with a random permutation series of itself. Now, each row contains a different parameter set that is used for each simulation run.

>>>>

\newcommand{\etal}{et al.}
\newcommand{\shir}{Shirreff et al.}%{\citeauthor{shirreff_transmission_2011}}
\newcommand{\champr}{Champredon et al.}%{\citeauthor{champredon_hiv_2013}}
\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
\newcommand{\rzero}{{\cal R}_0}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=small]{todonotes}

@@

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. The trade-off theory \citep{Ebert1999} --- which
postulates that parasite virulence can be explained as the long-term
evolutionary outcome of a saturating relationship between parasite
clearance rate and transmission rate --- has been criticized
\citep{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in a variety of host-pathogen systems \citep{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
particularly interesting application of these ideas is the work by
Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in a study of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
within-couple transmission
\citep{Fraser+2007,fraser_virulence_2014}\todo{read/check fraser+
  2014!; check vs. \cite{herbeck_is_2012}}. Subsequent studies
\citep{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately [XXX] years and then declining
slightly to a long-stable virulence level.

@@

\todo{more papers on SPVL heritability (\emph{maybe} include): \cite{alizon_phylogenetic_2010,leventhal_potential_2016}}

@@

@@

In contrast, models 5 and 6 are not explicitly structured.  Model 5 is an implicit serial monogamy model based on the epidemiological model used by \shir. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \citep{hollingsworth_hiv1_2008}. Finally, model 6 is a simple random-mixing model.

@@

\begin{equation}
z_i = \text{exp}(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
\end{equation}
For simplicity, we assume that all parameters follow logarithmic distribution. Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$). Then, each column is replaced with a random permutation series of itself. Now, each row contains a different parameter set that is used for each simulation run.
